---
title: "Homework 02 Data Handling, Graphics, More R"
output:
  pdf_document: 
     fig_height: 5
     fig_width : 7
  word_document: 
     fig_height: 5
     fig_width : 7
  html_document: default
subtitle: Due by 11:59pm, Friday, 1.31.25
author: S&DS 230/530/ENV 757
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**(1) Obama Tweets: Retweets vs. Favorites** A `.CSV` file containing Tweets from former President Barack Obama can be downloaded [HERE](http://reuningscherer.net/S&DS230/data/ObamaTweetsNEW.csv). The data is sorted by date, most recent at the top.

The variables (columns) are:

-   `text`: the body of the tweet
-   `date`: when the tweet was sent, original format
-   `date2`: when the tweet was sent, JUST the day (not the time of day)
-   `retweet_count`: how many people retweeted this tweet
-   `favorite_count`: how many people favorited this tweet
-   `is_retweet`: whether or not this tweet is a retweet of someone else's tweet
-   `source`: device used to send the tweet
-   `is_quote`: is the tweet a quote of someone else
-   `is_reply`: is the tweet a reply to someone else
-   `possibly_sensitive`: does the tweet possibly contain sensitive material

You can read more about retweets vs. replies [HERE](https://www.diffen.com/difference/ReTweet_vs_Reply).

There are two ways in which other Twitter users can indicate support for a tweet: *favoriting* and *retweeting*. For example, if a tweet has `favorite_count` = 5 and `retweet_count` = 10, then this suggests that 5 people favorited the tweet (saved it) and 10 people retweeted it (broadcasted it to their followers).

(1.1) Insert an R code chunk right below this that imports the data into a dataframe called `recent`. Note that the data is sorted in reverse time order. Get the header names of `recent` to confirm that the data imported correctly. Look at the first few rows of the data and the final few rows of the data. Also get the dimension of `recent`. What is the date range of the tweets? How many tweets does this dataset include?

```{r}
recent <- read.csv('http://reuningscherer.net/S&DS230/data/ObamaTweetsNEW.csv')
recent$X <- NULL
dim(recent)
names(recent)
head(recent, 5)
tail(recent, 5)
```

*Since the dataframe has 2000 columns, this dataset contains 2000 tweets. The first column is a tweet from January 13th, 2023, and the last column is a tweet from April 8th, 2016, so that is the date range.*

(1.2) Create a table that shows how many of Tweets came from each source and call this object `table1`. Show the results of `table1`. Write a single line that calculates the proportion of Tweets that were from Twitter Web Client, rounds this value to two decimal places, multipies the results by 100, and pastes on a "%" symbol. There should be no space between the number and the '%' symbol.

```{r}
table1 <- table(recent$source)
table1
paste(round(nrow(recent[recent$source == "Twitter Web Client",]) 
            / nrow(recent), digits=2) * 100, "%", sep="")
```

(1.3) Create a barplot that shows the number of tweets from each source. The labels of the barplot should also contain the whole number percentages for each tweet source (i.e. Thunderclap (14%) as an example (this isn't the correct percentage). Take the time to format your graph, and make sure the bars are horizontal. You'll want to include the commented line of code below AND you'll want to use the barplot option `cex.names = .6`. Write a comment in your code that explaines what this option does.

```{r}
par(mar=c(4,8,1,1)) # change the margins to fit the horizontal labels
barplot1 <- barplot(table1, 
                    horiz=TRUE, # make bars horizontal
                    cex.names=.8, # change the font size/proportions
                    las=1, # make labels horizontal
                    main="Obama's Tweets by Source (4/2016 - 1/2023)", 
                    xlim=c(0, 800),
                    xlab="Tweet Count")
text(x=table1, 
     y=barplot1 + 0, 
     adj=-0.2,
     cex=.8,
     labels=paste0(round(proportions(table1), digits=2) * 100, "%"))
```

(1.4) Get summary statistics for both `favorite_count` and `retweet_count`. Make histograms for each of these two variables as well. Put a title on each histogram, label the horizontal axis, and make the bars red. How would you describe the shape of these distributions (use words like 'symmetric' or 'skewed', or perhaps the name of some distribution that has a similar shape . . .)?

```{r}
summary(recent$favorite_count)
summary(recent$retweet_count)
options(scipen=5)
hist(recent$favorite_count, 
     col='red', 
     lwd=2, 
     xlab='Favorite Count', 
     main="Favorite Count of Obama's Tweets",
     breaks=20)
hist(recent$retweet_count, 
     col='red', 
     lwd=2, 
     xlab='Retweet Count', 
     main="Retweet Count of Obama's Tweets",
     breaks=20)
```

*Both the distributions of the favorite counts and the retweet counts of Obama's tweets are right skewed and appear to follow a geometric distribution.*

(1.5) Get summary statistics for `retweet_count` FIRST for the observations for which `is_quote` is `TRUE`, then for the observations for which 'is_quote`is`FALSE\`. Compare the medians of these two distributions - what do you observe?

```{r}
summary(recent[recent["is_quote"] == TRUE,]$retweet_count)
summary(recent[recent["is_quote"] == FALSE,]$retweet_count)
```

*The median retweet count for Obama's tweets that are quotes is higher than the median retweet count for his tweets that are not quotes.*

(1.6) Create a new dataframe called `recent_NoQuote` that contains all data from `recent` for which `is_quote` is `FALSE` (essentially, we're removing quotes and only looking at strictly original texts). USE THIS NEW DATAFRAME for the remainder of this problem set. Get the dimension of this dataframe. Compare this to a table of `is_quote` for the entire dataset to make sure the remaining number of rows (and columns) is correct.

Finally, make two new variables as a part of `recent_NoQuote` which will be the log transformations of `favorite_count` and `retweet_count`. Call these variables `logfavCnt` and `logreCnt`, respectively. The function you want to take log is called `log()`.

```{r}
recent_NoQuote <- recent[recent["is_quote"] == FALSE,]
dim(recent_NoQuote)
table(recent$is_quote)
recent_NoQuote$logfavCnt <- log(recent_NoQuote$favorite_count)
recent_NoQuote$logreCnt <- log(recent_NoQuote$retweet_count)
head(recent_NoQuote, 10)
```

(1.7) Make histograms of these two new log-scale variables. Put a title on each histogram, label the horizontal axis, and make the bars green How would you describe the shape of these transformed distributions (use words like 'symmetric' or 'skewed')?

```{r}
hist(recent_NoQuote$logfavCnt, 
     col='green', 
     lwd=2, 
     xlab='Log Favorite Count', 
     main="Logarithmic Favorite Count of Obama's Tweets",
     breaks=20)
hist(recent_NoQuote$logreCnt, 
     col='green', 
     lwd=2, 
     xlab='Log Retweet Count', 
     main="Logarithmic Retweet Count of Obama's Tweets",
     breaks=20)
```

*Both of the logarithmic distributions are unimodal and right skewed.*

(1.8) Make a plot of the number of times that each tweet was favorited vs. the number of times a tweet was retweeted. Put `favorite_count` on the y-axis and `retweet_count` on the x-axis. Label your axes, put on a main title, and make the plot characters blue.

```{r}
plot(recent_NoQuote$retweet_count, 
     recent_NoQuote$favorite_count, 
     xlab="Retweet Count", 
     ylab="Favorite Count", 
     main="Favorite Count vs. Retweet Count for Obama's Tweets", 
     col="Blue")
```

(1.9) Repeat part (1.8) but use the log-transformed variables. Label your axes, put on a main title, and make the plot characters red. How does the scatterplot on the log-scale compare to the scatterplot on the raw scale? Which one do you prefer?

```{r}
plot(recent_NoQuote$logreCnt, 
     recent_NoQuote$logfavCnt, 
     xlab="Log Retweet Count", 
     ylab="Log Favorite Count", 
     main="Log Favorite Count vs. Log Retweet Count for Obama's Tweets", 
     col="red")
```

*The data points are more evenly distributed across the entire scatterplot, which makes it easier to see an underlying trend compared to the previous scatterplot where the majority of data points were amassed in one area.*

(1.10) Create two new variables on the `recent_NoQuote` dataframe called `year` and `month` that will contain respectively the year and month the tweet was created. You'll need to look up how to use the function `substr()`. You'll also need to use the `as.numeric()` function to make sure that both new variables are numbers. Show the first 20 observations for each resulting variable.

```{r}
recent_NoQuote$year <- as.numeric(substr(recent_NoQuote$date2, 1, 4))
recent_NoQuote$month <- as.numeric(substr(recent_NoQuote$date2, 6, 7))
head(recent_NoQuote$year, 20)
head(recent_NoQuote$month, 20)
```

(1.11) Repeat part (1.9) BUT only for 2019 and 2020. First, create a dataframe called `recent_3` that only has observations from the specified years. You might want to use the `%in%` operator on your newly created variable year. Use this new dataframe to make your plot. Use the graphics option `pch = 19` to get solid round points, and make sure you have different colors for each of the two years. Finally, make sure your plot has a legend.

```{r}
recent_3 <- recent_NoQuote[recent_NoQuote$year %in% c(2019, 2020),]
plot(recent_3$logreCnt, 
     recent_3$logfavCnt, 
     xlab="Log Retweet Count", 
     ylab="Log Favorite Count", 
     main="Log Favorite Count vs. Log Retweet Count 
     for Obama's Tweets in 2019 and 2020",
     pch=19,
     col=ifelse(recent_3$year == 2019, "red", "blue"))
legend(x = "topleft", 
       legend = c(2019, 2020),
       col = c("red","blue"),
       pch = 19)
```

(1.12) Write no more than three sentences that describe what you see. Does the pattern appear any different between 2019 and 2020?

*For the most part, the relationship between log retweet count and log favorite count seems to be the same between 2019 and 2020. 2020 seems to have more variation with the outliers being farther out than in 2019.*
