---
title: "Homework 01 R Basics"
output:
  pdf_document: 
     fig_height: 5
     fig_width : 7
  word_document: default
  html_document: default
subtitle: Due by 11:59pm, Friday, 1.24.25
author: S&DS 230/530/ENV 757
---

------------------------------------------------------------------------

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**(1) RMarkdown Practice** *(24 points)* Change the markdown code below as indicated.

**Make this line bold**

*Make this line italics*

## Make this line a second level header

-   Make this line a bullet point
    -   Make this line an indented (or level two) bullet point

[**LINK**](https://www.nytimes.com/) (make the word LINK at left link to the New York Times home page AND make it bold)

``` r
Make this line look like R Code
```

Below this line, insert a new R chunk, create a vector called `xvec` that contains the integers 5 through 9, and have R display what is in `xvec`.

```{r}
xvec <- 5:9
xvec
```

**(2) R Syntax Practice** *(12 points)* Modify the R code below to follow good R Syntax practices

```{r, eval = FALSE}
x <- 5

x <- c(1, 2, 3)

length(x)

for (i in 1:10) {
  x <- 1+1
}

x <- 1
y <- c(3, 4)

```

**(3) Data handling** *36 pts*

(3.1) Insert a new R code chunk below.

```{r}
wb <- read.csv("http://reuningscherer.net/S&DS230/data/WB_2024.csv")
dim(wb)
names(wb)
head(wb, 5)
sapply(wb, class)
# class(wb[, "GNI"])
wb_Subset <- wb[wb$GNI > 70000 & is.na(wb$GNI)==F,
                c("Country", "GNI", "EnergyUse", "Measles")]
wb_Subset
wb_Stats <- summary(wb["Debt"], digits = 2)
wb_Stats
length(wb_Stats)
wb_Stats[c(1, 2, 3, 5, 6)]
```

(3.2) Read the .csv stored [HERE](http://reuningscherer.net/S&DS230/data/WB_2024.csv) into a new data frame and call is "wb". This is similar to the World Bank data I discussed in class two (this is a more current version). You can get the longer description of each variable [HERE](http://reuningscherer.net/S&DS230/data/WB_2024_NMS.csv)

(3.3) Get the dimension of wb.

(3.4) Get the variable names of wb.

(3.5) Show the first 5 lines of wb.

(3.6) Get the data type of each variable.

(3.7) In your code, insert a comment that gives the data type of the variable `GNI`?

(3.8) Create a new object called `wb_Subset` that has only the variables Country, GNI, EnergyUse, and Measles (in that order) AND only for countries where GNI is greater than 70000. Make sure you show the value of `wb_Subset`.

(3.9) Get summary statistics for `Debt`. Store the results in a new object called `wb_Stats`. Incidentally, `wb_Stats` will be a vector!

(3.10) Get the length of `wb_Stats`.

(3.11) Get r to show the following elements of `wb_Stats` : 1,2,3,5,6 AND round the result to 1 decimal place.

**(4) Plots** *16 pts*

(4.1) Using the WB dataset loaded above, make a scatterplot of "Rural" on the x axis and "Measles" on the y axis. Include a main title, axis titles, and a non-default symbol color and symbol type. *Hint: check out ?par or see examples from class 1 or class 3*.

(4.2) Make a boxplot of the the following `wb` variables that are all percentages: `Measles`, `Military`, `Renewable`, `Rural`. You want all four variables on one plot. Ensure the plot has a main title, axis labels, and a unique color for each variable.

```{r}
plot(wb$Rural, wb$Measles, 
     xlab="Rural", 
     ylab="Measles", 
     main="Scatterplot of Rural vs Measles", 
     col="Blue", pch='x')
boxplot(wb$Measles, wb$Military, wb$Rural, wb$Renewable, 
        names=c("Measles", "Military", "Renewable", "Rural"), 
        main="Boxplot of Various Country Facts",
        ylab="Percentage",
        col=c("Orange", "Red", "Green", "Brown"), 
        lwd=2)
```

**(5) Lists** *12 pts* The code below creates a list called `aList`

(5.1) Compute the sum of the second element of the list's third element. Store the result into an object named `mySum`. You'll want to use the `sum()` function.

```{r}
aList <- list(c(1, 5, 4), letters[c(1, 6, 4, 9, 22, 3)], list(c(1, 1, 1), 
                        c(14, 13, 12), c(3, 2, 1)), c(runif(8)))
mySum <- sum(aList[[3]][[2]])
mySum
```

(5.2) What is the difference between what is returned from the following two commands?

```{r }

aList[[3]][2]
aList[[3]][[2]]
```

*Both commands have the same first part, aList[[3]], which extracts the third element in aList, which is list(c(1, 1, 1), c(14, 13, 12), c(3, 2, 1)). The first command, which has [2], returns the second element of this list contained in a 1-element list, or list(c(14, 13, 12)). The second command, which has [[2]], just returns the second element of the list, which is c(14, 13, 12).*

 \
     
