---
title: "Homework 01 R Basics"
output:
  pdf_document: 
     fig_height: 5
     fig_width : 7
  word_document: default
  html_document: default
subtitle: Due by 11:59pm, Friday, 1.24.25
author: S&DS 230/530/ENV 757
---

------------------------------------------------------------------------

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**(1) RMarkdown Practice** *(24 points)* Change the markdown code below as indicated.

**Make this line bold**

*Make this line italics*

## Make this line a second level header

-   Make this line a bullet point
    -   Make this line an indented (or level two) bullet point

[**LINK**](https://www.nytimes.com/) (make the word LINK at left link to the New York Times home page AND make it bold)

`Make this line look like R Code`

Below this line, insert a new R chunk, create a vector called `xvec` that contains the integers 5 through 9, and have R display what is in `xvec`.

```{r}
xvec <- 5:9
xvec
```

**(2) R Syntax Practice** *(12 points)* Modify the R code below to follow good R Syntax practices

```{r, eval = FALSE}
x = 5

x<=c(1, 2, 3)

length (x)

for (i in 1:10) { 
x <-1+1 
           }

x <- 1 ; y <- c(3, 4)

```

**(3) Data handling** *36 pts*

(3.1) Insert a new R code chunk below.

```{r}

```

(3.2) Read the .csv stored [HERE](http://reuningscherer.net/S&DS230/data/WB_2024.csv) into a new data frame and call is "wb". This is similar to the World Bank data I discussed in class two (this is a more current version). You can get the longer description of each variable [HERE](http://reuningscherer.net/S&DS230/data/WB_2024_NMS.csv)

(3.3) Get the dimension of wb.

(3.4) Get the variable names of wb.

(3.5) Show the first 5 lines of wb.

(3.6) Get the data type of each variable.

(3.7) In your code, insert a comment that gives the data type of the variable `GNI`?

(3.8) Create a new object called `wb_Subset` that has only the variables Country, GNI, EnergyUse, and Measles (in that order) AND only for countries where GNI is greater than 70000. Make sure you show the value of `wb_Subset`.

(3.9) Get summary statistics for `Debt`. Store the results in a new object called `wb_Stats`. Incidentally, `wb_Stats` will be a vector!

(3.10) Get the length of `wb_Stats`.

(3.11) Get r to show the following elements of `wb_Stats` : 1,2,3,5,6 AND round the result to 1 decimal place.

**(4) Plots** *16 pts*

(4.1) Using the WB dataset loaded above, make a scatterplot of "Rural" on the x axis and "Measles" on the y axis. Include a main title, axis titles, and a non-default symbol color and symbol type. *Hint: check out ?par or see examples from class 1 or class 3*.

(4.2) Make a boxplot of the the following `wb` variables that are all percentages: `Measles`, `Military`, `Renewable`, `Rural`. You want all four variables on one plot. Ensure the plot has a main title, axis labels, and a unique color for each variable.

**(5) Lists** *12 pts* The code below creates a list called `aList`

(5.1) Compute the sum of the second element of the list's third element. Store the result into an object named `mySum`. You'll want to use the `sum()` function.

```{r}
aList <- list(c(1, 5, 4), letters[c(1, 6, 4, 9, 22, 3)], list(c(1, 1, 1), 
                        c(14, 13, 12), c(3, 2, 1)), c(runif(8)))
```

(5.2) What is the difference between what is returned from the following two commands?

```{r }

aList[[3]][2]
aList[[3]][[2]]
```

*replace this text with your answer AND your answer should be in italics when knitted*

 \
     
